Liveness-Based Garbage Collection for Lazy Languages

### 2015-04-27 paper/paper.tex
To dos for Prasanna

\documentclass[10pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage {algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

  We circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
  strongly  regular}  CFGs which  have   easy
translations to non-deterministic finite state automata (NFA). The
NFAs are then simplified on the
basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
membership of forward access paths. The resulting NFAs are   finally
converted to DFAs for use during garbage collection.
\comment{Fix all comments and produce first draft}
%------------------------------------------------------------%
\subsubsection{Approximating CFGs using NFAs.}
\label{sec:NFA-approx}

We use the   
algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
approximate a CFG  to a   {\em strongly regular\/} grammar.  The
transformation has  the  property that  if $\Lv$  is a
non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
  L}_{G'}(\Lv)$.    This  is   required  for the
approximation to be safe with respect to liveness.

@@

%------------------------------------------------------------%
\subsubsection{Transforming NFAs to Accept Forward Paths:}
%\label{sec:NFA-elim}
The strongly regular CFGs obtained after the Mohri-Nederhof
transformation are
  first converted  into  NFAs.  The        algorithm         described        in
Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow} \lang{\nt}$, where {\nt} accepts forward paths only.
Thus  \nt\  can  be  used  to  check  
membership of forward paths.

@@

The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
process is  continued till  a fixed point  is reached.   When the
fixed point  is reached, the resulting NFA  contains all possible
reductions corresponding  to all the  paths in the  original NFA.
The proofs  of the termination  and correctness of  the algorithm
are given in our earlier paper~\cite{karkare07liveness}.

\input{liveness-dfa.tex}

We  illustrate  the  algorithm  in  Figure~\ref{algo:simplify-NFA}  by
constructing      the       automaton      for      \var{\Lanv{9}{y}}.
Figure~\ref{fig:automata-construction}  (c)  shows  the automaton  for
\var{\Lanv{9}{y}}   constructed   by   composing  the   automata   for
\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
get (d). We  add an $\epsilon$ edge from $q_2$  to $q_3$ bypassing the
\bcdr\acdr\ pair from $q_2$ to $q_2$  and then to $q_3$. This is shown
in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
$q_2$.   Removing this $\epsilon$  edge gives  the automaton  shown in
(g).  Restricting  this automaton  to forward edges  only, we  get the
final   automaton   shown    in   (h).   This   automaton   recognizes
$\acdr^{*}\!\mid\acdr^{*}\acar  \sigma_{\mathit \!all}$,  showing that
the entire list \py, including  its elements, is live at $\pi_9$. Also
note that the  language accepted by the final  automaton satisfies the
prefix-closed property.

@@

\section{Implementation}
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.  An important difference in  GC of lazy language compared to
an  eager language is  the point  of garbage  collection.  In  case of
eager  languages only  \CONS~ points  and  function calls  are the  GC
points. In  a lazy  language every \LET~  statement is a  potential GC
point. Hence we need to store liveness automata corresponding to every
\LET~  statement in the  program. This  leads to  an explosion  in the
number of  automata. From our  observation that along an  {\em E-path}
all program points have the same liveness, we can reduce the number of
automata state required  by sharing the same automata  for all program
points along an {\em E-path}. In  the example program, we only need to
create automata  for program  points \ldots, since  they are  the only
distinct {\em E-path}s in the program.
% Abstract description of the GC algorithm
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
%% \STATE $new\_car\gets$ copy($cell.car$)
%% \STATE $new\_cell.car\gets$ $new\_car$ 
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ELSE
%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
%% \ELSE
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \ENDIF
%% \ENDIF
%% \end{algorithmic}

% Liveness for cons cells & reachability for closures
During GC, if the heap cell being processed is a cons cell, we consult
the  liveness automaton  corresponding to  it and  copy only  the live
paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
approximate the  liveness to be {\{0+1\}\*} and  copy everything under
the closure.
% Handling of mixture of LGC & RGC
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closure or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a later GC
% and copy wrong locations. Avoiding it by setting non-live references to null after every GC.
% Proof for the correctness. 
% Difference between eager and lazy GC

Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ENDIF
%% \end{algorithmic}
We   provide   an   intuition    about   the   correctness   of   this
algorithm. Consider the following facts about the print stack during a
GC,

>>>>

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage {algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

  We circumvent the problem by over-approximating the CFG generated by the analysis to  {\em
  strongly  regular}  CFGs which  have   easy
translations to non-deterministic finite state automata (NFA). The
NFAs are then simplified on the
basis  of  the  $\hookrightarrow$  rules to  enable  checking  of
membership of forward access paths. The resulting NFAs are   finally
converted to DFAs for use during garbage collection.
\comment{Fix all comments and produce first draft}
%------------------------------------------------------------%
\subsubsection{Approximating CFGs using NFAs.}
\label{sec:NFA-approx}
\comment{Change the language and the example}
We use the   
algorithm  by  Mohri  and Nederhof~\cite{mohri00regular}   to
approximate a CFG  to a   {\em strongly regular\/} grammar.  The
transformation has  the  property that  if $\Lv$  is a
non-terminal  in the grammar  $G$ and  $G'$ is  the grammar  after the
Mohri-Nederhof transformation, then  ${\mathscr L}_G(\Lv) \subseteq {\mathscr
  L}_{G'}(\Lv)$.    This  is   required  for the
approximation to be safe with respect to liveness.

@@

%------------------------------------------------------------%
\subsubsection{Transforming NFAs to Accept Forward Paths:}
%\label{sec:NFA-elim}
The strongly regular CFGs obtained after the Mohri-Nederhof
transformation are
  first converted  into  NFAs.  The        algorithm         described        in
Figure~\ref{algo:simplify-NFA} converts  an NFA \ntbar\  to a NFA
\nt\ such that  $\lang{\ntbar} \stackrel{*}{\hookrightarrow}
\lang{\nt}$, where {\nt} accepts forward paths only.
\comment{Rewrite the algorithm to handle \cbar and update the example automata}
Thus  \nt\  can  be  used  to  check  
membership of forward paths.

@@

The algorithm repeatedly introduces  $\epsilon$ edges to bypass a
pair of consecutive edges labelled \bcar\acar\ or \bcdr\acdr.  The
process is  continued till  a fixed point  is reached.   When the
fixed point  is reached, the resulting NFA  contains all possible
reductions corresponding  to all the  paths in the  original NFA.
The proofs  of the termination  and correctness of  the algorithm
are given in an earlier paper~\cite{karkare07liveness}.

\input{liveness-dfa.tex}
\comment{change according to example}
We  illustrate  the  algorithm  in  Figure~\ref{algo:simplify-NFA}  by
constructing      the       automaton      for      \var{\Lanv{9}{y}}.
Figure~\ref{fig:automata-construction}  (c)  shows  the automaton  for
\var{\Lanv{9}{y}}   constructed   by   composing  the   automata   for
\var{\Uf{\append}{1}}, \var{\Df{\append}{1}} and $(\epsilon \mid \acdr
\mid \acdr\acar\sigma_{\mathit{\!all}})$.  After $\epsilon$ removal we
get (d). We  add an $\epsilon$ edge from $q_2$  to $q_3$ bypassing the
\bcdr\acdr\ pair from $q_2$ to $q_2$  and then to $q_3$. This is shown
in (e).  The $\epsilon$ edge is removed in (f) and a second $\epsilon$
is added to the automaton bypassing the \bcar\acar\ pair from $q_1$ to
$q_2$.   Removing this $\epsilon$  edge gives  the automaton  shown in
(g).  Restricting  this automaton  to forward edges  only, we  get the
final   automaton   shown    in   (h).   This   automaton   recognizes
$\acdr^{*}\!\mid\acdr^{*}\acar  \sigma_{\mathit \!all}$,  showing that
the entire list \py, including  its elements, is live at $\pi_9$. Also
note that the  language accepted by the final  automaton satisfies the
prefix-closed property.

@@

\section{Implementation}
By the  time the program  starts executing, the liveness  automata are
created and stored in a file.   This is re-used every time the program
executes.  An important difference in  GC of lazy language compared to
an  eager language is  the point  of garbage  collection.  In  case of
eager  languages only  \CONS~ points  and  function calls  are the  GC
points. In  a lazy  language every \LET~  statement is a  potential GC
point. Hence we need to store liveness automata corresponding to every
\LET~  statement in the  program. This  leads to  an explosion  in the
number of  automata. \comment{Directly say that we share the automata}From our  observation that along an  {\em E-path}
all program points have the same liveness, we can reduce the number of
automata state required  by sharing the same automata  for all program
points along an {\em E-path}. In  the example program, we only need to
create automata  for program  points \ldots, since  they are  the only
distinct {\em E-path}s in the program.
% Abstract description of the GC algorithm
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(copied\_using\_rgc)$}
%% \STATE $new\_car\gets$ copy($cell.car$)
%% \STATE $new\_cell.car\gets$ $new\_car$ 
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ELSE
%% \IF {$celltype\ is\ closure\ and\ cell\ not\ already\ copied$}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \STATE $new\_cell.arg1\gets$ copy($cell.arg1$)  
%% \STATE $new\_cell.arg2\gets$ copy($cell.arg2$)  
%% \ELSE
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \ENDIF
%% \ENDIF
%% \end{algorithmic}

% Liveness for cons cells & reachability for closures

\comment{Mention the fact that liveness at a program point indicates
  what references will be used by the program eventually. The exact
  point where a reference will be used cannot be determined statically
due to the language being lazy. }
During GC, if the heap cell being processed is a cons cell, we consult
the  liveness automaton  corresponding to  it and  copy only  the live
paths.  In  case  the  heap  cell  points  to  a  closure,  we  safely
approximate the  liveness to be {\{0+1\}\*} and  copy everything under
the closure.
% Handling of mixture of LGC & RGC
A consequence of  copying everything under a closure  is the fact that
we can end up trying to copy things which are not live and corrupt the
heap. This  can happen if at some  $GC_k$ a cons cell  is copied using
liveness. At  a later stage this  cons cell becomes part  of a closure
and  hence during $GC_{k+n}$  the algorithm  tries to  copy everything
under the  cons cell  thus copying heap  locations which might  not be
valid.  To avoid  this behavior we either have to choose  to do a full
liveness based  GC for closure or  we should ensure  that any non-live
references are  not carried across  GC's. We take the  latter approach
and ensure that  non-live references are set to  null. After every GC,
the live part of the buffer  is scanned and any reference which is not
live is  set to null.  The correctness of  this follows from  the fact
that any reference which is decalred  not live in a $GC_k$ will not be
derefernced during later execution.
% A reference copied using LGC could become part of a closure in a later GC
% and copy wrong locations. Avoiding it by setting non-live references to null after every GC.
% Proof for the correctness. 
% Difference between eager and lazy GC

Another  difference that is  present in  GC of  lazy language  is the
presence  of  a print  stack.  In  lazy  languages the  evaluation  of
closures is  driven by the print  function\comment{Refer to SPJ's book}.  Hence, during  a GC there
will be heap  references that will be present on  the stack which also
have  to  copied/collected.   One  solution  is  to  assume  that  all
references  present  on  the  print  stack are  fully  live  and  copy
everything pointed  by them  akin to reachability.   This is  safe but
copies heap  cells which  have already been  printed. Since  the print
function is  not part of  the executing program its  liveness automata
are not  considered. Therefore we  simulate the liveness of  the print
function by using a flag in  the heap cell. The flag ensures that only
the required parts  of a cons cell are copied. The  behavior of the GC
algorithm is as described below,
%% \begin{algorithmic}
%% \STATE $new\_cell\gets$ dup\_heapcell(cell)
%% \IF{$celltype\ is\ cons\ and\ not(processed\_car\_part)$}
%% \STATE $new\_cdr\gets$ copy(cell.cdr)
%% \STATE $new\_cell.cdr\gets$ $new\_cdr$
%% \ENDIF
%% \end{algorithmic}
We   provide   an   intuition    about   the   correctness   of   this
algorithm. Consider the following facts about the print stack during a
GC,
